---
title: "just_census_data"
author: "Fiona Edwardson"
date: "3/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidycensus)
library(sf)
library(tigris)
library(rmapshaper)
library(tmap)
library(areal)
library(leaflet)
library(dbplyr)
# install.packages("censusapi")
library(censusapi)
```

```{r putting in Census API key}

census_api_key("MY_KEY", install = TRUE, overwrite = TRUE)

Sys.getenv("CENSUS_API_KEY")

readRenviron("~/.Renviron")
```


# data set 1: total population & white population

```{r setting wd}
# setting wd
setwd("~/Desktop/GISII files/FinalProject")

```

```{r importing the Demographic data (pop data)}

# Just checking -- it's the same data! the api request worked! 

# head(demo2019)
# head(subset(demo2019[ , c(1, 130:150)]))

demog2019 <- get_acs(geography = "tract", 
              year = 2019,
              variables = c(tpop = "DP05_0001E", twhite = "DP05_0037E"),
              state = "IL",
              survey = "acs5",
              output = "wide",
              geometry = TRUE)

view(demog2019)

demog2015 <- get_acs(geography = "tract", 
              year = 2015,
              variables = c(tpop = "DP05_0001E", twhite = "DP05_0032E"),
              state = "IL",
              survey = "acs5",
              output = "wide",
              geometry = TRUE)

view(demog2015)

```

DP05_0001E == total population
DP05_0037E == total population > one race > white (column ~ 133)
+ need GEOID

## r data wrangling
```{r subsetting only the tracts in cook county}

demog19 <-
  demog2019 %>%
  filter(stringr::str_detect(NAME, 'Cook County') )

view(demog19)


demog15 <-
  demog2015 %>%
  filter(stringr::str_detect(NAME, 'Cook County') )

glimpse(demog15)

```


```{r making a new column -- % white}

demg19 <- mutate(demog19, pwhite = twhite/tpop)
demg15 <- mutate(demog15, pwhite = twhite/tpop)
glimpse(demg15)

```

```{r subsetting: only % white and geoid}

dem19 <- subset(demg19, select = c(GEOID, pwhite) )
dem15 <- subset(demg15, select = c(GEOID, pwhite) )

glimpse(dem15)
```


# data set 2: % poverty

```{r pulling in poverty data}

pov2015 <- get_acs(geography = "tract", 
              year = 2015,
              variables = c(tpov = "S1701_C02_001E"),
              state = "IL",
              county = "cook",
              survey = "acs5",
              output = "wide",
              show_call = TRUE,
              geometry = TRUE)

glimpse(pov2015)

plot(pov2015$tpov)

summary(pov2015)

pov2019 <- get_acs(geography = "tract", 
              year = 2019,
              variables = c(tpov = "S1701_C02_001E"),
              state = "IL",
              county = "Cook",
              survey = "acs5",
              output = "wide",
              geometry = TRUE)
```



```{r getting tot pop as its own data set}

totpop19 <- subset(demg19, select = c(GEOID, tpop), left = TRUE )
totpop15 <- subset(demg15, select = c(GEOID, tpop), left = TRUE )

df15 <- data.frame("GEOID" = totpop15$GEOID, "tpop" = totpop15$tpop)
glimpse(df15)

df19 <- data.frame("GEOID" = totpop19$GEOID, "tpop" = totpop19$tpop)

```

Actually, let me do the same for poverty:
```{r pov as a data set (w/o geometry col)}
tpov15 <- data.frame("GEOID" = pov2015$GEOID, "tpov" = pov2015$tpov)
tpov19 <- data.frame("GEOID" = pov2019$GEOID, "tpov" = pov2019$tpov)

glimpse(tpov19)

```


now merging the data:

```{r merging totpop to poverty}

pov15 <- left_join(df15, tpov15, by = "GEOID")
pov19 <- left_join(df19, tpov19, by = "GEOID")

glimpse(pov15)

```

Seems a little fishy, at first glance. However, let us soldier on. (Note: I was using the wrong variable key !)

```{r making a percent pov category}

pov19 <- mutate(pov19, ppov = tpov/tpop)
pov15 <- mutate(pov15, ppov = tpov/tpop)
glimpse(pov15)

plot(pov15["ppov"])

```

Nice!


## data set 3: median household income

Hopefully I can just pull this one in:

```{r pulling in the data}
h_income2015 <- get_acs(geography = "tract", 
              year = 2015,
              variables = c(med_earn = "S2001_C01_002E"),
              state = "IL",
              county = "cook",
              survey = "acs5",
              output = "wide",
              show_call = TRUE,
              geometry = TRUE)

glimpse(h_income2015)

# plot(h_income2015$tpov)

summary(pov2015)

h_income2019 <- get_acs(geography = "tract", 
              year = 2019,
              variables = c(med_earn = "S2001_C01_002E"),
              state = "IL",
              county = "Cook",
              survey = "acs5",
              output = "wide",
              geometry = TRUE)

glimpse(h_income2019)
```

Took longer than ideal to figure out what the variable name for the median earnings was called (the census data site is really not very well set up).


```{r subsetting for final df}

incdf15 <- data.frame("GEOID" = h_income2015$GEOID, "medearn" = h_income2015$med_earn)

incdf19 <- data.frame("GEOID" = h_income2019$GEOID, "medearn" = h_income2019$med_earn)

glimpse(incdf15)

```


## data set 4: median house value

```{r pulling in the data}
houseval2015 <- get_acs(geography = "tract", 
              year = 2015,
              variables = c(med_house = "DP04_0089E"),      
              #Estimate!!VALUE!!Owner-occupied units!!Median (dollars)
              state = "IL",
              county = "cook",
              survey = "acs5",
              output = "wide",
              show_call = TRUE,
              geometry = TRUE)

glimpse(houseval2015)

# plot(h_income2015$tpov)

houseval2019 <- get_acs(geography = "tract", 
              year = 2019,
              variables = c(med_house = "DP04_0089E"),
              state = "IL",
              county = "Cook",
              survey = "acs5",
              output = "wide",
              geometry = TRUE)

glimpse(houseval2019)
```

```{r makinguseful dfs}

dfhval15 <- data.frame("GEOID" = houseval2015$GEOID, "medHouse" = houseval2015$med_house)

dfhval19 <- data.frame("GEOID" = houseval2019$GEOID, "medHouse" = houseval2019$med_house)

glimpse(dfhval19)
```


## final data tables

We'll conclude with 2 separate tables, one for the index for 2015, one for the index for 2019.

```{r merging final dfs}

dem15 <- data.frame("GEOID" = dem15$GEOID, "pwhite" = dem15$pwhite)
dem19 <- data.frame("GEOID" = dem19$GEOID, "pwhite" = dem19$pwhite)

glimpse(pov15)
fdf15 <- left_join(pov15, dem15, by = "GEOID") 
fdf15 <- left_join(fdf15, incdf15, by = "GEOID")
fdf15 <- left_join(fdf15, dfhval15, by = "GEOID")

glimpse(fdf15)

fdf19 <- left_join(pov19, dem19, by = "GEOID") 
fdf19 <- left_join(fdf19, incdf19, by = "GEOID")
fdf19 <- left_join(fdf19, dfhval19, by = "GEOID")

glimpse(fdf19)

```

I need to do some final edits:
```{r choosing final variables}

ffdf15 <- data.frame("GEOID" = fdf15$GEOID, "pwhite" = fdf15$pwhite, "ppov" = fdf15$ppov, "medearn" = fdf15$medearn, "medHouse" = fdf15$medHouse)

glimpse(ffdf15)

ffdf19 <- data.frame("GEOID" = fdf19$GEOID, "pwhite" = fdf19$pwhite, "ppov" = fdf19$ppov, "medearn" = fdf19$medearn, "medHouse" = fdf19$medHouse)

glimpse(ffdf19)

```



## writing the dfs

I'm first going to make the folders in the directory that I want.

```{r saving the census data}
getwd()

write_csv(ffdf15, "Census_Data_For_Index/2015/Cen2015.csv")
write_csv(ffdf19, "Census_Data_For_Index/2019/Cen2019.csv")
```


# Making the Index:

```{r summary (find missing values)}

summary(ffdf15)

```

Ok, so we'll need to use na.rm = TRUE in some of our later operations. Let's ignore that for now.

## Standardizing variables

In order to combine variables into an index, we need to standardize them. We do this to get the variables onto the same scale. A common approach to standardizing a variable is to calculate its z-score. The z-score is a measure of distance from the mean, in this case the mean of all tracts in an area. So, after standardizing, the variables will have the same units, specifically units of standard deviations.

```{r calculating z scores}

ffdf15 %>%
  mutate(pwhite = (pwhite-mean(pwhite, na.rm=TRUE))/sd(pwhite, na.rm=TRUE)) %>%
  select(GEOID,pwhite) %>%
  summarize(Mean = mean(pwhite, na.rm = TRUE), SD = sd(pwhite, na.rm = TRUE)) 



```

```{r attempting to do this for all variables in the df}

cen15.std <- ffdf15 %>%
        mutate_at(~(scale(.) %>% as.vector(.)), .vars = vars(-c(GEOID)))

summary(cen15.std)

```

It seems to have worked, although I didn't take out the NA values! Yay!

Repeating for the 2019 data:

```{r 2019 index data}

cen19.std <- ffdf19 %>%
        mutate_at(~(scale(.) %>% as.vector(.)), .vars = vars(-c(GEOID)))

summary(cen19.std)

```

Moving forward, I'm just going to take the average of the variables. I'm going to focus on the census data for 2015 at the moment:

```{r adding avg col}

cen15.std <- cen15.std %>%
  mutate(DepInd15 = (pwhite+ppov+medearn+medHouse)/4)

glimpse(cen15.std)

cen19.std <- cen19.std %>%
  mutate(DepInd19 = (pwhite+ppov+medearn+medHouse)/4)

glimpse(cen19.std)
```

```{r selecting only index values}
Idf15 <- data.frame("GEOID" = cen15.std$GEOID, "I15" = cen15.std$DepInd15)
Idf19 <- data.frame("GEOID" = cen19.std$GEOID, "I19" = cen19.std$DepInd19)

Idf <- left_join(Idf15, Idf19, by = "GEOID")

print(Idf)

```

```{r putting in a change factor}

Idf <- mutate(Idf, Ichange = I19 - I15)

Idf <- data.frame("GEOID" = Idf$GEOID, "IDel" = Idf$Ichange)
plot(Idf$IDel)

```

Index done!

Writing the data as a csv file:

```{r}
write_csv(Idf, "Census_Data_For_Index/IndexVals.csv")
```

